Cross-scale integration of knowledge for predicting species ranges: a metamodeling framework

### 2015-08-28 tex/appendix_s1.tex
Nearly final draft

Because Phenofit provided predictions in the form of a probability of presence (rather than directly in terms of the metamodel parameters as in the previous example), it was necessary to relate these predictions to the metamodel.
We opted to do this via simulation because it is a natural method to use in an MCMC scheme.
We thus treated the Phenofit predictions (\(\psi_P\)) as an observed result of a true underlying set of occurrences (\(X_P\)).
Because these occurrences were unobserved (and in the case of the future distribution, unobservable), it was necessary to infer them using MCMC.
For each MCMC iteration, we generated a simulated occurrence dataset \(\hat{X}_P\) by drawing from a Binomial distribution with probabilities equal to \(\psi_P\).
In order to give both the occurrence dataset and Phenofit predictions equal weight, we drew the same number of samples as were present in the occurrence dataset.
The metamodel was then fit to these simulated data (with conditioning on the naive model as well):
\begin{equation} 
\overbrace{p(\theta_M \mid X_M, D_M, X_P, \psi_P)}^\text{integrated posterior}
\propto
\overbrace{p\left (X_P \mid \theta_M,D_M,\psi_P \right )}^{\substack{\text{new information} \\ \text{from Phenofit}}}
\overbrace{p \left(X_M \mid \theta_M, D_M \right) p \left(\theta_M \right)}^{\substack{\text{naive metamodel} \\ \text{posterior}}}
\overbrace{p \left(\psi_P \right)}^{\substack{\text{prior for} \\ \text{Phenofit}}}	
\end{equation}
This procedure, by generating a random dataset with each iteration, incorporated the variance in the Phenofit predictions.
Finally, because we were interested in two separate questions with respect to Phenofit integration (one regarding how integration affects uncertainty in our present estimate of the species distribution and a one regarding the effect of integration on projections into the future), we performed two separate integrations with the Phenofit information.
To address the first question, we used the Phenofit predictions for the present as \(\psi_P\) and present climate as \(D_M\), producing the Integrated-Present model.
To address the second, we used the future predictions from Phenofit and future climate, resulting in the Integrated-Future model.
In both cases, the naive model was a GLM relating present occurrences to present climate (and thus the prior response to the environment was static, similarly to how one might use the same conventional SDM to both model present distribution and project into the future).
Convergence for all models was assessed by comparing the results of multiple chains with overdispersed starting points.
Based on the results of these initial runs, we thinned all posterior samples by recording only every 50th sample, and we obtained 25000 posterior samples for each model following a burn-in period of 20000 samples.
Finally, we evaluated the Naive and Integrated-Present models by computing the area under the receiver operating curve (AUC).
AUC measures the ability of the model to discriminate presences and absences, where a value of 1 indicates perfect discrimination and 0.5 is random discrimination.
Because we obtained joint posterior distributions for all models, it is not possible to evaluate the performance of a `best' model; rather, we computed AUC for every posterior sample of both models, resulting in posterior distributions of AUC (Fig. \ref{fig:auc}).
Median AUC values were 0.802 and 0.797 for the Naive and Integrated-Present models, respectively, and the overlap between the two distributions suggests little difference in the performance of each model.

@@

>>>>

Because Phenofit provided predictions in the form of a probability of presence (rather than directly in terms of the metamodel parameters as in the previous example), it was necessary to relate these predictions to the metamodel.
We opted to do this via simulation because it is a natural method to use in an MCMC scheme.
We thus treated the Phenofit predictions (\(\psi_P\)) as an observed result of a true underlying set of occurrences (\(X_P\)).
Because these occurrences were unobserved (and in the case of the future distribution, unobservable), it was necessary to infer them using MCMC.
For each MCMC iteration, we generated a simulated occurrence dataset \(\hat{X}_P\) by drawing from a Binomial distribution with probabilities equal to \(\psi_P\).
In order to give both the occurrence dataset and Phenofit predictions equal weight, we drew the same number of samples as were present in the occurrence dataset.
The metamodel was then fit to these simulated data (with conditioning on the naive model as well):
\begin{equation} 
\overbrace{p(\theta_M \mid X_M, D_M, X_P, \psi_P)}^\text{integrated posterior}
\propto
\overbrace{p\left (X_P \mid \theta_M,D_M,\psi_P \right )}^{\substack{\text{new information} \\ \text{from Phenofit}}}
\overbrace{p \left(X_M \mid \theta_M, D_M \right) p \left(\theta_M \right)}^{\substack{\text{naive metamodel} \\ \text{posterior}}}
\overbrace{p \left(\psi_P \right)}^{\substack{\text{prior for} \\ \text{Phenofit}}}	
\end{equation}
This procedure, by generating a random dataset with each iteration, incorporated the variance in the Phenofit predictions.
Finally, because we were interested in two separate questions with respect to Phenofit integration (one regarding how integration affects uncertainty in our present estimate of the species distribution and a one regarding the effect of integration on projections into the future), we performed two separate integrations with the Phenofit information.
To address the first question, we used the Phenofit predictions for the present as \(\psi_P\) and present climate as \(D_M\), producing the Integrated-Present model.
To address the second, we used the future predictions from Phenofit and future climate, resulting in the Integrated-Future model.
In both cases, the naive model was a GLM relating present occurrences to present climate (and thus the prior response to the environment was static, similarly to how one might use the same conventional SDM to both model present distribution and project into the future).
Convergence for all models was assessed by comparing the results of multiple chains with overdispersed starting points.
Based on the results of these initial runs, we thinned all posterior samples by recording only every 50th sample, and we obtained 25000 posterior samples for each model following a burn-in period of 20000 samples.
Finally, we evaluated the Naive and Integrated-Present models by computing the area under the receiver operating curve (AUC) and by computing calibration curves using reserved data.
AUC measures the ability of the model to discriminate presences and absences, where a value of 1 indicates perfect discrimination and 0.5 is random discrimination.
Because we obtained joint posterior distributions for all models, it is not possible to evaluate the performance of a `best' model; rather, we computed AUC for every posterior sample of both models, resulting in posterior distributions of AUC (Fig. \ref{fig:auc}).
Median AUC values were 0.802 and 0.797 for the Naive and Integrated-Present models, respectively, and the overlap between the two distributions suggests little difference in the performance of each model.
Calibration curves suggested that the Integrated-Present model overpredicted somewhat compared to the Naive model (Fig. \ref{fig:calib}).
This is unsurprising, as the validation data were a reserved subset of the same data used to fit the naive model, while the Integrated-Present model included additional information.
Furthermore, fitted values from Phenofit were generally greater in the same geographic region than the Naive model (see Fig. 6, main manuscript), explaining the overprediction of the Integrated-Present model and suggesting that values from the two models may not be exactly numerically equivalent, despite the interpretation of both as a probability of presence.
No model evaluation was performed for the Integrated-Future model, because no independent dataset is available for future predictions.

@@

\begin{figure}
\includegraphics[width=6.5in]{figs/ex2_calib.png}
\caption{Calibration curve for the Naive (left panel) and Integrated-Present (right panel) models.
The calibration curve compares the predicted probability of presence from the model (horizontal axis) to the observed proportion of presences (left axis); values were computed by classifying grid cells into 20 equal-width bins and performing a linear regression (weighted by the sample size in each bin) of predicted values against observed frequencies.
Axes end at 0.7 because neither model predicted probabilities of presence greater than this value.
Solid lines are posterior mean values, shaded regions are 95\% credible intervals.
The dashed line indicates perfect calibration.}
\label{fig:calib}
\end{figure}
